---
title: "Homework 6"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1. Warm up: Log-Odds, Odds and Probability <small>5 points; 1 point each</small>

a. Convert a log-odds of 1.75 to probability.

***

Approx 0.852

***

b. Convert probability of 0.85 to odds.

***

Approx. 5.67

***

c. Event A has a probability of 0.7 and the odds ratio of A to B is 1.45. Calculate the probability of event B.

***

Approx 0.617

***

d. You roll a single 6 sided die. What are the odds you get a number divisible by 3?

***

Approximately 0.5

***

e. The odds ratio comparing event A to B is 9 and the risk ratio is 3. What is $Pr(A)$? *Hint: let $Pr(A)=a$ and $Pr(B)=b$, and write out the odds ratio and risk ratio as equations. 

***

0.75

***



## Problem 2. Interpreting logistic regression <small>6pts; 2 pts each</small>

Suppose we collect data for a group of students in a statistics class with independent variables $X_{1}=\text{hours studied}$, $X_{2}=\text{GPA}$, and binary response variable
$$
Y= \begin{cases} 1 &\mbox{ if student received an A} \\
  0 &\mbox{ otherwise. }
  \end{cases}
$$
Suppose that we fit a logistic regression model to the data, predicting $Y$ from $X_1$ and $X_2$ (and an intercept term) and produce estimated coefficients $\hat{\beta}_{0}=-6, \hat{\beta}_{1}=0.05, \hat{\beta}_{2}=1$.

### Part a) Logistic regression and probability

According to our fitted model, what is the probability that a student receives an A if they study for $40$ hours and have a GPA of $3.5$?

```{r}

# Given coefficients
beta0 <- -6
beta1 <- 0.05
beta2 <- 1

# Given values for X1 and X2
X1 <- 40
X2 <- 3.5

# Calculate the logit
logit <- beta0 + beta1 * X1 + beta2 * X2

# Calculate the probability
probability <- exp(logit) / (1 + exp(logit))
probability

```

### Part b) Interpreting coefficients
According to our fitted model, an additional hour spent studying is associated with *how much* of an increase in the log odds of receiving an A?

```{r}

beta1

```

### Part c) "Inverting" logistic regression probabilities
According to our fitted model, how many hours would the student in Part (a) need to study to have a $50\%$ chance of getting an A in the class?
That is, keeping GPA fixed at $3.5$, how many hours of study are needed so that the probability of an A is $50\%$?
If you aren't up for the math, feel free to find an approximate solution via guess-and-check in R.

***

We need to find how many hours \( X1 \) a student needs to study to have a 50% chance of getting an A, keeping the GPA fixed at 3.5. At 50% probability, the log odds \( \log\left(\frac{p}{1-p}\right) \) is 0.

Set up the equation:
\[ 0 = -6 + 0.05X1 + 1 \times 3.5 \]
\[ 0 = -6 + 0.05X1 + 3.5 \]
\[ 0.05X1 = 2.5 \]
\[ X1 = \frac{2.5}{0.05} \]

***

```{r}
# Solve for X1
X1_needed <- (2.5 - 3.5) / 0.05
X1_needed

```



## Problem 3. Palmer Penguins Part I  <small>9pts; 3 pts each</small>

The Palmer Penguin dataset (https://allisonhorst.github.io/palmerpenguins/) consists of 344 observations of penguins belonging to 3 penguin species across the islands in the Palmer Archipeligo in Antarctica. We will build a logistic model attempting to classify the penguins based on physical characteristics. For each penguin we've recorded: 

* `species` - the species, either "Gentoo", "Adelie" or "Chinstrap"
* `island` - which of three islands the Penguin was observed (Biscoe, Dream or Torgersen)
* `bill_length_mm` - the length of the bill in mm
* `bill_depth_mm` - the depth of the bill (vertical thickness of the closed bill)
* `filler_length_mm` - length of their cute flippers
* `body_mass_g` - the body mass in grams
* `sex` - female, male or NA (unknown)
* `year` - The year of the observation: 2007, 2008 or 2009

First you need to download the library. Run this chunk of code once.
```{r, eval=FALSE, echo=FALSE}
#Run this code once to install the library
install.packages("palmerpenguins")
```

Then load the library and the penguin dataset. Note: Your RMD won't knit until you run the above chunk.
```{r}
library(palmerpenguins)
```

### a) Adelie penguins based on island

We are going to try to classify penguins as **Adelie** or **not Adelie**. So create a new variable called `Adelie` which will be 1 or 0 based on whether the penguin species is Adelie

```{r}
data("penguins")

# Create the binary variable for Adelie species
penguins$Adelie <- as.integer(penguins$species == "Adelie")
```

Perform some analysis looking at each of the 3 islands - create a 2 way table between `island` and the `Adelie` variable. Look at the proportions conditioned on island. What proportion of observations on each island were Adelie?

```{r}
table_island_adelie <- table(penguins$island, penguins$Adelie)

prop_adelie_by_island <- prop.table(table_island_adelie, 1)
prop_adelie_by_island
```

***
It appears that Torgersen only has Adelie penguins, whereas Biscoe and Dream only have 26.2% and 45.2% respectively.

***

### b) Adelie Penguins on Dream

Find the (i) probability, (ii) odds and (iii) log-odds that a randomly selected penguin from Dream is an Adelie penguin?

```{r}
dream_data <- penguins[penguins$island == "Dream",]

prob_adelie_dream <- mean(dream_data$Adelie)

odds_adelie_dream <- prob_adelie_dream / (1 - prob_adelie_dream)

log_odds_adelie_dream <- log(odds_adelie_dream)

prob_adelie_dream
odds_adelie_dream
log_odds_adelie_dream
```

***
(i) 0.452
(ii) 0.824
(iii) -0.194

***

### c) An island-based classifier

Now fit a logistic model predicting whether a penguin is Adelie based on island.
Interpret the intercept and the coefficient of the `islandDream` variable. Use this model to predict the probability that a penguin from Dream is Adelie.

```{r}
# Fit logistic model
model <- glm(Adelie ~ island, family = "binomial", data = penguins)

# Summary to view coefficients
summary(model)

# Predict probability for a penguin from Dream island
predicted_prob <- predict(model, data.frame(island = "Dream"), type = "response")
predicted_prob
```

***
the probability that a penguin from Dream is Adelie is 0.452

***



## Problem 4. Penguins Part II <small>10pts; 2 pts each</small>

In this problem we will work once again with the Palmer Penguin dataset. We will work with a subset by taking out all missing values. After you have installed the package and loaded the library, uncomment the line below.
```{r}
penguins.complete <- penguins[complete.cases(penguins),]
```

### a) Predicting Palmer Penguins with quantitative Predictors

Now use the two bill measurements (`bill_length_mm` and `bill_depth_mm` as predictors in a new logistic model. Suppose a penguin with a bill length of 53.1 and a bill depth of 22.7 is observed. What is the model's probability that the penguin is an Adelie penguin?

```{r}
model <- glm(species == "Adelie" ~ bill_length_mm + bill_depth_mm, family = "binomial", data = penguins.complete)

predicted_probability <- predict(model, newdata = data.frame(bill_length_mm = 53.1, bill_depth_mm = 22.7), type = "response")
predicted_probability
```

***

Prob ~ 0.079

***

### b) Interpreting coefficients

Are longer bills associated with an increased or decreased likelihood that a penguin is an Adelie penguin?

```{r}
summary(model)
```


***

- The coefficient for `bill_length_mm` is -2.2099, with a p-value of 0.00125, which is statistically significant at typical alpha levels (like 0.01 or 0.05). This indicates that as the bill length increases, the log-odds of a penguin being an Adelie decrease. In practical terms, this means longer bills are associated with a decreased likelihood of a penguin being an Adelie.

- The coefficient for `bill_depth_mm` is 3.9981, with a p-value of 0.00705, which is also statistically significant. This suggests that as the bill depth increases, the log-odds of a penguin being an Adelie increase. Therefore, deeper bills are associated with an increased likelihood of a penguin being an Adelie.

These interpretations are consistent with the signs of the coefficients: a negative sign for bill length indicates an inverse relationship with being an Adelie, while a positive sign for bill depth indicates a direct relationship.

***

### c) A full classifier

Fit a logistic model to predict whether a penguin is a **Chinstrap** penguin using all four of the biological measurements (`bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g`). 

```{r}
# Fit logistic model for Chinstrap penguins
chinstrap_model <- glm(species == "Chinstrap" ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g, family = "binomial", data = penguins.complete)

# Summary to view coefficients and significance
summary(chinstrap_model)
```

Which of the predictors are significant?

***

- The `bill_length_mm` coefficient is 2.774014, with a p-value of 0.00113. This is statistically significant, indicating that longer bills are associated with a higher likelihood of a penguin being a Chinstrap.
  
- The `bill_depth_mm` coefficient is 0.109140, with a p-value of 0.78114. This is not statistically significant, suggesting that bill depth does not have a reliable association with being a Chinstrap penguin in this model.
  
- The `flipper_length_mm` coefficient is -0.199360, with a p-value of 0.06220. This p-value is marginal; it suggests that there might be an association where penguins with longer flippers are less likely to be Chinstrap, but the evidence is not strong enough to assert this conclusively at the typical alpha level of 0.05.
  
- The `body_mass_g` coefficient is -0.015230, with a p-value of 0.00129. This is statistically significant, indicating that heavier penguins are less likely to be Chinstrap.

In conclusion, bill length and body mass are the significant predictors for identifying a Chinstrap penguin in this model. Longer bill lengths increase the likelihood, whereas higher body mass decreases the likelihood. Bill depth is not a significant predictor, and flipper length is borderline significant and would require further study or data to confirm its effect.

***


### d) Assessing the model
Suppose you will predict that a penguin is a Chinstrap if the estimated $\hat{y}=\sigma(\hat{z})>0.5$. When predicting using this threshold, what is the type 1 error rate of your predictor on the dataset? What is power of the predictor on the dataset? 
*Hint: you will want to compare the predicted $\hat{y}$ values to the actual $y$ values. The `table` command can produce a 2x2 confusion matrix to help you answer this question.*


```{r}
# Predictions at the default threshold of 0.5
predictions <- predict(chinstrap_model, type = "response") > 0.5

# Actual values
actuals <- penguins.complete$species == "Chinstrap"

# Create a confusion matrix
conf_matrix <- table(Predicted = predictions, Actual = actuals)

# Calculate type 1 error rate and power
type1_error_rate <- conf_matrix["TRUE", "FALSE"] / sum(actuals == "FALSE")
power <- conf_matrix["TRUE", "TRUE"] / sum(actuals == "TRUE")
type1_error_rate
power
```

***

alpha = 0.003773585
power = 0.9558824

***


### e) Adjusting the Type 1 error rate

Now modify your threshold from 0.5 to some other threshold with the goal of achieving the highest power possible while keeping the type 1 error rate  below 0.05. What threshold would you use? What is the type 1 error rate and power of this new classifier?

```{r}
# Define a function to calculate type 1 error and power for various thresholds
calculate_metrics <- function(threshold) {
  predictions <- predict(chinstrap_model, type = "response") > threshold
  conf_matrix <- table(Predicted = predictions, Actual = actuals)
  type1_error <- conf_matrix["TRUE", "FALSE"] / sum(actuals == "FALSE")
  power <- conf_matrix["TRUE", "TRUE"] / sum(actuals == "TRUE")
  return(c(Type1Error = type1_error, Power = power))
}

# Test various thresholds to find the best balance
sapply(seq(0.1, 0.9, by = 0.01), calculate_metrics)
```

***

The threshold that maintains a type 1 error rate below 0.05 is at the threshold of approximately 0.75 (as seen in the output where the type 1 error rate switches from 0.0037385 to 0), and the corresponding power at this threshold is approximately 0.897 (or 89.7%).

Therefore, if you want to keep the type 1 error rate below 0.05 and achieve as high power as possible, you would use a threshold of about 0.75, which gives you a type 1 error rate of approximately 0 and a power of approximately 89.7%.

***
