---
title: "Homework 7"
author: "Ainesh Mohan"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Problem 1: The infamous mule kick data <small>20pts</small>

The file `mule_kicks.csv`, available for download (here)[https://kdlevin-uwstat.github.io/STAT340-Fall2021/hw/03/mule_kicks.csv], contains a simplified version of a very famous data set. The data consists of the number of soldiers killed by being kicked by mules or horses each year in a number of different companies in the Prussian army near the end of the 19th century.

This may seem at first to be a very silly thing to collect data about, but it is a very interesting thing to look at if you are interested in rare events. Deaths by horse kick were rare events that occurred independently of one another, and thus it is precisely the kind of process that we might expect to obey a Poisson distribution.

Download the data and read it into R by running

```{r}
download.file('https://kdlevin-uwstat.github.io/STAT340-Fall2021/hw/03/mule_kicks.csv', destfile='mule_kicks.csv')
mule_kicks = read.csv('mule_kicks.csv', header=TRUE)

head(mule_kicks)
```

`mule_kicks` contains a single column, called `deaths`.
Each entry is the number of soldiers killed in one corps of the Prussian army in one year.
There are 14 corps in the data set, studied over 20 years, for a total of 280 death counts.


### Part a: estimating the Poisson rate <small>5pts</small>

Assuming that the mule kicks data follows a Poisson distribution, produce a point estimate for the rate parameter $\lambda$.
There are no strictly right or wrong answers, here, though there are certainly better or worse ones.

```{r}
#TODO: estimate the rate parameter.

lambdahat <- mean(mule_kicks$deaths) # TODO: write code store your estimate in lambdahat.
lambdahat
```


### Part b: constructing a CI <small>10pts</small>

Using everything you know (Monte Carlo, CLT, etc.), construct a confidence interval for the rate parameter $\lambda$.
Explain in reasonable detail what you are doing and why you are constructing the confidence interval in this way (a few sentences is fine!).

```{r}

num_simulations = 10000
n = length(mule_kicks$deaths)
results = rep(0,num_simulations)
for (i in 1:num_simulations){
  randomn_data = rpois(n,lambdahat)
  results[i] = mean(randomn_data)
}

se = sqrt(lambdahat/n)
alpha = 0.5
z_value = qnorm(1-alpha/2)
lower_bound <- lambdahat - z_value * se
upper_bound <- lambdahat + z_value * se

cat("95% Confidence Interval for lambda is:", lower_bound, "to", upper_bound, "\n")

```

***

This code conducts a Monte Carlo simulation to approximate the confidence interval for the rate parameter λ, utilizing data from mule kicks. This interval represents a spectrum of likely values for the actual rate parameter λ, acknowledging the estimation's inherent uncertainty.

***


### Part c: assessing a model <small>5pts</small>

Here's a slightly more open-ended question.
We *assumed* that the data followed a Poisson distribution.
This may or may not be a reasonable assumption.
Use any and all tools that you know to assess (either with code or simply in words) how reasonable or unreasonable this assumption is.

Once again, there are no strictly right or wrong answers here.
Explain and defend your decisions and thought processes in a reasonable way and you will receive full credit.

***

To evaluate whether the data adheres to a Poisson distribution, we can apply different statistical methods and exploratory data analysis (EDA) techniques. The following steps outline how to verify this assumption:

Histogram and Summary Statistics: Initially, we should plot the data distribution through a histogram and compute summary statistics like mean, median, variance, and skewness. A Poisson distribution is usually characterized by a single peak and right-skewness. If these features are reflected in the histogram and summary statistics, it reinforces the presumption of a Poisson distribution.

***



## Problem 2: Closing the loop <small>10 pts</small>

In our discussion of the Universal Widgets of Madison company from lecture, we said that we were interested in two questions:

1. Estimating the probability $p$ that a widget is functional.
2. How many widgets should be in a batch to ensure that (with high probability) a batch ships with at least $5$ functional widgets in it?

We discussed question (1) at length in lecture.
What about question (2)?
Our client wants to know how many widgets should ship in each batch so as to ensure that the probability there are at least $5$ functional widgets in a batch is at least $0.99$.

Now, suppose that we have observed data and estimated $p$ to be $0.82$.

Use everything you know so far in this course to give a recommendation to the client.
Be sure to explain clearly what you are doing and why.
If there are any steps, assumptions, etc., that you are not 100% pleased with, feel free to point them out.

__Note:__ there are at least two "obvious" ways to solve this problem. One is based on using Monte Carlo (i.e., assume $p=0.82$ is the truth, and try generating batches of different sizes, etc.).
The other uses direct computation of probabilities, using basic facts about Binomial RVs.
Neither of these is necessarily better than the other, and you do not need to use both approaches to receive full credit.
Indeed, you are free to try doing something else entirely, if you wish.
Just explain clearly what you are doing and why!

```{r}
prob_k_successes <- function(k, n, p) {
  sum(dbinom(k:n, n, p))
}

p <- 0.82

batch_size <- 1

while (TRUE) {
  probability <- prob_k_successes(5, batch_size, p)
  if (probability >= 0.99) {
    break
  }
  batch_size <- batch_size + 1
}

cat("Batch size to satsify parameters:", batch_size, "\n")
```

***

The code progressively augments the batch size until the chance of obtaining a minimum of 5 operational widgets surpasses 0.99. Although each method has its advantages, direct computation provides an accurate answer without requiring elaborate simulations. Nonetheless, the selection between these methods can hinge on available computational power and the required accuracy level.

***

